---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# lwo

<!-- badges: start -->
<!-- badges: end -->

The goal of lwo is to estimate Longitudinal Win Odds (lwo) from longitudinal extension of the Probabilistic Index Model (PIM)

## Installation

You can install the development version of lwo from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("Yongxi-Long/lwo")
```

## Example

The key function is *lwo()*, which fits a longitudinal probabilistic index model and estimate longitudinal win odds for ordinal repeated measurements. 

We use example data set *dat_SID* from this package for illustration. It is a simulated data set derived from the SID-GBS trial (Second intravenous immunoglobulin dose in patients with Guillain-Barré syndrome with poor prognosis (SID-GBS): a double-blind, randomised, placebo-controlled trial
Walgaard, ChristaWalgaard, Christa et al.
The Lancet Neurology). *dat_SID* contains both the wide format data.frame and the long format data.frame, of the following variables:

-   patient_ID: unique patient identifier
-   age: patient age at entry
-   pre_diarrhea: whether had preceding diarrhea before entry
-   treat_ITT: treatment assignment under intention-to-treat
-   GBS_DS: Guillain-Barré syndrome disability scale at week 0, 1, 2, 4, 8, 12, 26

### The *lwo()* function

```{r}
library(lwo)
# load the data set
data("dat_SID")
# extract the long format
dat_SID_long <- dat_SID$long_format
# fit a longitudinal PIM model
# we estimate the treatment effect quantified as win odds,
# adjusted for age, preceding diarrhea and baseline GBS-DS score
# we also model treatment-by-time interaction via a spline term
mod.lwo.spline <- lwo(
  GBS_DS ~ treat_ITT + age + pre_diarrhea + GBS_DS_baseline+
    treat_ITT:splines::ns(week,knots = 2),
  data = dat_SID_long,
  id.var = "patient_ID",
  visit.var = "week",
  time.vars = "week",
  corstr = "ar1",
  larger = FALSE
)
summary(mod.lwo.spline)

newdata <- dat_SID_long |>
  dplyr::filter(patient_ID %in% c(1,4))
predict(object = mod.lwo.spline,
        newdata = newdata,
        id.var = "patient_ID",
        visit.var = "week",
        time.vars = "week",
        type = "link",
        conf.int = TRUE) |>
  exp() |>
  data.frame()
```

### The *gen_data()* function

The package also provides a function, the *gen_data()*, to simulate ordinal longitudinal data based on user inputs. The function internally relies on the *genOrdCat()* function from the *simstudy* package.

```{r}
# define the covariate distribution
# if covariate_def is left NULL, the function will use build-in distributions for (continuous) age variable and (binary) preceding diarrhea variable.
def <- simstudy::defData(varname = "sex", dist = "binary", formula = 0.5)
def <- simstudy::defData(def, varname = "age50", dist = "normal", formula = 0, variance = 100)
# define baseline category probabilities when all covariates are zero/reference values
baseprobs <- c(0.3,0.4,0.3)
# correlation matrix, must match the dimension of the visits
corMatrix <- gen_corMatrix(n_visits = 4,rho=0.6,corstr = "ar1")
dat <- gen_data(
   N = 200,
   baseprobs = baseprobs,
   covs_effects = c(sex = 0.5, age50 = -0.02),
   time_effects = c(0.1, 0.3, 0.5),
   trt_ratio = 1,
   time_trt_effects = c(0.05, 0.1, 0.15),
   visits = paste0("v", 1:4),
   corMatrix = corMatrix,
   covariate_def = def
 )
```

### The *calculate_win_odds()* function

The longitudinal ordinal data is generated from a proportional odds model, in which the effects are specified as log odds ratio. As a results, it is not straightforward to see what the true win odds is for each visit after specifying the data generation paramters. This package provides a function, the *calculate_win_odds()*, that use simulation to compute the true (population-level) win odds at each visit, marginalized with respect to the covariate distribution.

The input of this function is similar to that of the *gen_data()* function, except that we have *N_approx* instead of *N* to specify the number of iterations of the simulation to approximate the true win odds.

```{r}
corMatrix <- gen_corMatrix(n_visits = 3,rho=0.6,corstr = "ar1")
estimands <- calculate_win_odds(N_approx = 1e4,
                               baseprobs = rev(c(0.06,0.11,0.12,0.50,0.21)),
                               covs_effects = c("age"=-0.005,
                                                "pre_diarrhea"= 0.23),
                               time_effects = c(0.6,1.2),
                               trt_ratio = 1,
                               time_trt_effects = c(0.4,0.8),
                               visits = visits <- c("week0","week4","week8"),
                               corMatrix = corMatrix)
estimands
```

